---
title: "Environmental Justice Index Exploration - West Texas"
author: "Grace Beasley"
date: "`r Sys.Date()`"
output:
  html_document:
    number_sections: yes
    toc: yes
    toc_float: yes
---

```{r setup, include=FALSE}

knitr::opts_chunk$set(
  echo = F, warning = F, message = F
  )

```


# Introduction

The Environmental Justice Index pulls information from the U.S. Census Bureau, EPA, CDC, and Mine Safety and Health Administration to create measures of factors associated with environmental health and ranks census tracts on cumulative environmental health justice measures. The cumulative rankings take into account social vulnerabilities, environmental burden, and health vulnerabilities. Social vulnerabilities include elements socioeconomic status, racial/ethnic minority status, household characteristics, and housing type. Environmental burden, which we will be focusing on in this report, include measures of air pollution, potentially hazardous and toxic sites, built environment, and water pollution. The health vulnerabilities focus primarily on pre-existing chronic burden of disease.

This analysis will conduct a linear regression to explore the association between average annual days above PM2.5 regulatory standards and the average health score of the county. The geographic distribution of average health scores will also be examined cartographically.

Because of the massive scope of this dataset, we have subset the geographic location to include only the census tracts that make up the counties of West Texas.

Variables of interest are defined as follows

Outcomes:

  - **F_HVM**: Sum of 5 binary health outcomes variables that designate a census tract as flagged (var = 1) if the tract is ranked above 0.6666 percentile in risk for the following: high blood pressure, asthma, cancer, mental health, and diabetes.

Exposure:

  - **E_PM**: Annual average days above PM2.5 regulatory standards.
  
Confounders:

  - **EP_POV200**: Percentage below 200% poverty
  - **EP_MINTRY**: Percentage of minoritized persons


## Objectives

1. To explore the association between average annual days over PM2.5 regulatory standards and health outcomes in West Texas
2. To visualize the aggregated mean health risk scores at the county level


```{r, load-data}
here::i_am("EJI_report_gnb.Rmd")
data <- readRDS(
  file = here::here("derived_data/data_west_aspatial.rds")
)
```


# Analysis

## Descriptive Figures

```{r figure}

figure1 <- readRDS(file = here::here("output/figure1.rds"))
figure1

```

As we can see, the highest mean health score reported was a 3, reported by Coke, Irion, Kimble, Mason, McCulloch, Menard, and Terrell County. The next highest mean score was a 2 while the lowest mean score was 0.25.

```{r figure 2}
map_path <- here::here("output/flags_inset.png")
```

![](`r map_path`)

For context, the values in the bar plot have been mapped. We see that the counties with the highest mean score appear to predominantly fall in the southern part of the region. For context, those counties sit south of the I-20 corridor and are remote and generally sparsely populated. There are a number of factors in such a scenario that can result in greater prevalence of adverse health outcomes including lack of access to care and higher rates of poverty. This clustering to the south suggests there may be geographically important factors at play, though a much more robust analysis of this area would be needed to support that.

## Regression table

```{r regression}

reg_table <- readRDS(here::here("output/reg_table.rds"))
reg_table

```

In the linear model exploring the association between annual days above PM2.5 regulatory standards and health scores, there does not appear to be a significant association after controlling for the percentage of population below 200% poverty and the percentage of minoritized persons. As mentioned above, this region of Texas is incredibly rural and as such, there are a number of factors outside of environmental exposure that can contribute to above average prevalences of adverse health outcomes. More nuanced investigations of health and environmental exposure in the general population of this region is warranted.
